---
title: "R Notebook"
output: html_notebook
---

```{r setup}
knitr::opts_knit$set(root.dir = "~/git/HushToFhir/")
```

This is the dataset for the phenotyping algorithms deposited at PHeKB.
Data retrived Aug 15, 2018

```{r}
library(readxl)
library(tidyverse)
```

load data
```{r}
phekb <- read_xlsx("./data/PheKB_phenotype_algorithms.xlsx", skip = 1, col_names = TRUE, sheet = 1)
```

Count the percentage of algorithms that used lab tests
```{r}
colnames(phekb) <- make.names(colnames(phekb))
methods <- phekb$Data.Modalities.and.Methods.Used.
all <- vector(mode = "character")
for (i in 1:length(methods)) {
  if (length(methods[i]) == 0 | is.na(methods[i])) {
    next
  }
  elements <- strsplit(methods[i], ",")[[1]]
  for (j in 1: length(elements)){
    e = trimws(elements[j]) # the trim is not working because, in short, there are different codings for space. 
    e = gsub("(^\\s+)|(\\s+$)", "", e)
    if (!is.element(e, all)){
      all[length(all) + 1] = e
    }
  }
}
print(all)
```

```{r}
names_all = make.names(all)
n = ncol(phekb)
for(i in 1:length(all)) {
  phekb[,n+i] = ifelse(str_detect(phekb$Data.Modalities.and.Methods.Used., all[i]), 1, 0)
}
colnames(phekb)[(n+1):(n+length(names_all))] = names_all
```

Algorithms that used laboratories
```{r}
counts = lapply(phekb[,(n+1):(n+length(names_all))], sum, na.rm = TRUE)
COUNT_ALL_METHODS = nrow(phekb)
methods_list = names(counts)
alg_count = unlist(counts)
data_alg_method_count <- data.frame(method = methods_list, count = alg_count)

# combine ICD9 and ICD10 counts
data_alg_method_count <- data_alg_method_count %>% 
  filter(str_detect(method, "ICD")) %>% 
  summarise(method = "ICD.Codes", count = sum(count)) %>%
  union(
    data_alg_method_count %>% filter(!str_detect(method, "ICD"))
  )

ggplot(data_alg_method_count) + geom_bar(aes(fct_reorder(method, count, .desc = TRUE), count/COUNT_ALL_METHODS), stat = "identity", fill = "black") + 
  xlab("Method Utilized") + ylab("percent of algorithms") +
  scale_y_continuous(labels = scales::percent) +
  theme_bw() + theme(panel.grid = element_blank(), axis.text.x = element_text(angle = 45, vjust = 0.5))
```

Conclusion:

In EMERGE phenotyping algorithms, lab test is only the fifth most frequently used data, behind diagnosis codes, medication, clinical notes and CPT procedure codes
